A geographic routing approach for IPv6 in large-scale low-power and lossy networks

### 2015-04-06 2014_ELSEVIER_CEE_SI/paper_cee_special_issue.tex
small fixes

\begin{abstract}
%% Text of abstract
In this paper, we propose \removed{a routing protocol for a smart street lighting system based on IPv6-enabled wireless sensor network} \rev{\textit{GeoRank}, a geographic routing approach for the IPv6-enabled large-scale  wireless networks}. We discuss the main drawbacks of the RPL (IPv6 Routing Protocol for
Low-Power and Lossy Networks) for P2P (point-to-point) communication in large-scale 6LoWPAN networks. Then, we address such drawbacks by proposing a \removed{new} routing \removed{algorithm} \rev{protocol}, named \textit{GeoRank}, which integrates RPL protocol with the position-based routing protocol GOAFR (Greedy Other Adaptive Face Routing). \removed{Simulations were performed on networks extracted from real street maps.} The \removed{obtained} results \rev{obtained with simulations} show that \textit{GeoRank} finds shorter routes than RPL in high link density conditions and than GOAFR in low link density conditions. Thus, \textit{GeoRank} shows to be adaptive to variable link densities found in large-scale networks. Further, \textit{GeoRank} avoids the use of DAO (Destination Advertisement Object) control messages required in RPL, while being more scalable in terms of memory usage than storing-mode RPL.
\end{abstract}

@@

However, RPL has some drawbacks, as higher amount of control messages than data messages, mainly for constructing routes from the sink to the nodes (downward routes in RPL vocabulary) \citep{Heurtefeux_RPL_2013,P2P_analysis_2010,P2P-RPL_2011}. \removed{In a street light system,} \rev{In a network where} nodes are static, \removed{and} a routing protocol could benefit of this condition to find routes with less control overhead than RPL. The routing algorithms that use position to find routes in a communication network are known as geographic routing algorithms and a large body of literature has been devoted to the research on those algorithms in last few decades \citep{Karp_GPSR_2000, GOAFR_2008, Denardin2011}.

Our main contribution in this paper is to propose a modification into RPL protocol, in order to better support P2P communication. The proposed technique intend to integrate a geographic routing algorithm in the RPL protocol to reduce the amount of control message required for P2P communication. \removed{We further propose that} Such a routing protocol can be used \rev{, for example,} in a smart street lighting system \removed{based on IPv6-enabled} wireless \removed{sensor} network for supporting diverse applications in an urban IoT. This paper is organized as follows: Section \ref{secRequirements} addresses the requirements of a wireless sensor network to implement an smart street lighting network. Also, we argue about the benefits of creating a network with such coverage for other public services. Section \ref{sec6LoWPAN} presents an overview of the 6LoWPAN protocol, describing succinctly its operating modes, routing strategies and main drawbacks. In section \ref{secGeographicRouting} we address the geographic routing technique, highlighting its pros and cons. In section \ref{secGeoRank}, the \textit{GeoRank} algorithm is presented, depicting its working principle. We evaluate, by simulation, the performance of the proposed algorithm in Section \ref{secSimulation}. Finally, we conclude this paper in Section \ref{secConclusion}.

@@

In the geographic routing approach, the destination address is the position of the destination node, instead of a topological address. To discover a node position in the network, each router node must implement a location service. This service is responsible to determine its own position and the destination position. A node position could be obtained by pre-configuration if the node is stationary, or by a GPS receiver. \removed{As} \rev{If} network nodes are static \removed{in the street lighting case}, the location information of the nodes can be stored during the installation process.

@@

It is remarkable that the \textit{GeoRank} is able to reduce the length of the route found by geographic routing in lower link densities. It does so by avoiding to circumvent the void with face routing whenever it is possible to use the DODAG to perform this deviation. Thus, \textit{GeoRank} achieves a significant reduction of the path length when the number of voids is large (i.e. when the radio range is small) compared to geographic routing alone. On the other hand, when the radio range is increased, geographic routing performs better than RPL, and \textit{GeoRank} does not impact negatively its performance. Therefore, \textit{GeoRank} works as a selective method, that is able leverage the best of RPL and geographic routing, and being adaptive to the network density.

In a large scale network \removed{, as a smart street lighting}, there may be some places where the link density is very low due to obstacles, such as in dense urban environments, while some places may have high link density due to less and lower buildings. Therefore, while in either case RPL or geographic routing alone would not be the best option, the remarkable adaptability of \textit{GeoRank} makes it the well suited to such variable link density case. Moreover, \textit{GeoRank} is more scalable than storing-mode RPL according to memory usage, as shown in figure \ref{sim_results2}.

@@

In this paper, it was proposed a new routing protocol, named \textit{GeoRank}, which integrates the standardized RPL routing protocol for 6LoWPAN networks with the state-of-the-art GOAFR position-based routing algorithm. 
Such a protocol is targeted to be used \removed{in a smart street lighting system based on} \rev{in a large-scale} IPv6-enabled wireless sensor network for supporting diverse applications in an urban IoT. The results obtained from simulations on networks extracted from real street maps have shown that \textit{GeoRank} has an improved routing performance over GOAFR in low link density condition, while outperforming the RPL protocol in high link density condition by finding mostly optimal routes. Therefore, \textit{GeoRank} is an adaptive approach for scenarios with variable link densities. Further, it is remarked that \textit{GeoRank} is not required to use the RPL DAO control messages, because it uses the positions of the nodes for the P2P routing, and is more scalable than storing-mode RPL according to memory usage.

>>>>

\begin{abstract}
%% Text of abstract
In this paper, we propose\removed{a routing protocol for a smart street lighting system based on IPv6-enabled wireless sensor network} \rev{\textit{GeoRank}, a geographic routing approach for the IPv6-enabled large-scale  low-power and lossy networks}. We discuss the main drawbacks of the RPL (IPv6 Routing Protocol for
Low-Power and Lossy Networks) for P2P (point-to-point) communication in large-scale 6LoWPAN networks. Then, we address such drawbacks by proposing a \removed{new}routing \removed{algorithm}\rev{protocol}, named \textit{GeoRank}, which integrates RPL protocol with the position-based routing protocol GOAFR (Greedy Other Adaptive Face Routing).\removed{Simulations were performed on networks extracted from real street maps.} The\removed{obtained} results \rev{obtained with simulations} show that \textit{GeoRank} finds shorter routes than RPL in high link density conditions and than GOAFR in low link density conditions. Thus, \textit{GeoRank} shows to be adaptive to variable link densities found in large-scale networks. Further, \textit{GeoRank} avoids the use of DAO (Destination Advertisement Object) control messages required in RPL, while being more scalable in terms of memory usage than storing-mode RPL.
\end{abstract}

@@

However, RPL has some drawbacks, as higher amount of control messages than data messages, mainly for constructing routes from the sink to the nodes (downward routes in RPL vocabulary) \citep{Heurtefeux_RPL_2013,P2P_analysis_2010,P2P-RPL_2011}. \removed{In a street light system,}\rev{In a network where} nodes are static, \removed{and}a routing protocol could benefit of this condition to find routes with less control overhead than RPL. The routing algorithms that use position to find routes in a communication network are known as geographic routing algorithms and a large body of literature has been devoted to the research on those algorithms in last few decades \citep{Karp_GPSR_2000, GOAFR_2008, Denardin2011}.

Our main contribution in this paper is to propose a modification into RPL protocol, in order to better support P2P communication. The proposed technique intend to integrate a geographic routing algorithm in the RPL protocol to reduce the amount of control message required for P2P communication.\removed{We further propose that} Such a routing protocol can be used\rev{, for example,} in a smart street lighting system \removed{based on IPv6-enabled} wireless \removed{sensor}network for supporting diverse applications in an urban IoT. This paper is organized as follows: Section \ref{secRequirements} addresses the requirements of a wireless sensor network to implement an smart street lighting network. Also, we argue about the benefits of creating a network with such coverage for other public services. Section \ref{sec6LoWPAN} presents an overview of the 6LoWPAN protocol, describing succinctly its operating modes, routing strategies and main drawbacks. In section \ref{secGeographicRouting} we address the geographic routing technique, highlighting its pros and cons. In section \ref{secGeoRank}, the \textit{GeoRank} algorithm is presented, depicting its working principle. We evaluate, by simulation, the performance of the proposed algorithm in Section \ref{secSimulation}. Finally, we conclude this paper in Section \ref{secConclusion}.

@@

In the geographic routing approach, the destination address is the position of the destination node, instead of a topological address. To discover a node position in the network, each router node must implement a location service. This service is responsible to determine its own position and the destination position. A node position could be obtained by pre-configuration if the node is stationary, or by a GPS receiver. \removed{As}\rev{If} network nodes are static\removed{in the street lighting case}, the location information of the nodes can be stored during the installation process.

@@

It is remarkable that the \textit{GeoRank} is able to reduce the length of the route found by geographic routing in lower link densities. It does so by avoiding to circumvent the void with face routing whenever it is possible to use the DODAG to perform this deviation. Thus, \textit{GeoRank} achieves a significant reduction of the path length when the number of voids is large (i.e. when the radio range is small) compared to geographic routing alone. On the other hand, when the radio range is increased, geographic routing performs better than RPL, and \textit{GeoRank} does not impact negatively its performance. Therefore, \textit{GeoRank} works as a selective method, that is able \rev{to} leverage the best of RPL and geographic routing, and being adaptive to the network density.

In a large scale network,\removed{as a smart street lighting,} there may be some places where the link density is very low due to obstacles, such as in dense urban environments, while some places may have high link density due to less and lower buildings. Therefore, while in either case RPL or geographic routing alone would not be the best option, the remarkable adaptability of \textit{GeoRank} makes it the well suited to such variable link density case. Moreover, \textit{GeoRank} is more scalable than storing-mode RPL according to memory usage, as shown in figure \ref{sim_results2}.

@@

In this paper, it was proposed a new routing protocol, named \textit{GeoRank}, which integrates the standardized RPL routing protocol for 6LoWPAN networks with the state-of-the-art GOAFR position-based routing algorithm. 
Such a protocol is targeted to be used \removed{in a smart street lighting system based on}\rev{in a large-scale} IPv6-enabled wireless sensor network for supporting diverse applications in an urban IoT. The results obtained from simulations on networks extracted from real street maps have shown that \textit{GeoRank} has an improved routing performance over GOAFR in low link density condition, while outperforming the RPL protocol in high link density condition by finding mostly optimal routes. Therefore, \textit{GeoRank} is an adaptive approach for scenarios with variable link densities. Further, it is remarked that \textit{GeoRank} is not required to use the RPL DAO control messages, because it uses the positions of the nodes for the P2P routing, and is more scalable than storing-mode RPL according to memory usage.
